---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r echo=FALSE, include=FALSE}
library(data.table)
library(magrittr)
```

# What cstidy is for

cstidy puts aggregated surveillance data into one standard shape: a
`csfmt_rts_data` table. You give it counts that already carry a time column and
a location column. It fills in the rest of the time and geography columns that
the shape requires, so everything downstream can assume those columns are there.

cstidy does not turn individual records into counts. Aggregate first, then hand
the result to cstidy.

# Which format to use

Use **`csfmt_rts_data_v3`**, through `set_csfmt_rts_data_v3()`. It is the
current format and what new work should target.

**`csfmt_rts_data_v1` and `csfmt_rts_data_v2` are deprecated.** The mark is a
signpost, not an alarm. Neither format prints a deprecation warning, and the
mark changed no executable line, so nothing you have running will break or
start complaining. The rest of this page documents v2 and is still accurate.
Do not start there.

The three formats are siblings. None of them inherits from another.

```{r}
d1 <- cstidy::generate_test_data()
d2 <- cstidy::generate_test_data()
d3 <- cstidy::generate_test_data()

cstidy::set_csfmt_rts_data_v1(d1)
cstidy::set_csfmt_rts_data_v2(d2)
cstidy::set_csfmt_rts_data_v3(d3)

class(d1)
class(d2)
class(d3)

inherits(d3, "csfmt_rts_data_v2")
```

# Three things to know before you move to v3

## v3 derives fewer columns, and removes none

Each format carries the list of columns it derives when the input lacks them.
v2's list holds 18 columns; v3's holds 11.

Deriving is not removing. `set_csfmt_rts_data_v3()` takes no column away, so a
table built under v2 keeps every column it had. What a move to v3 costs is the
guarantee that those columns are present, not the values in them.

```{r}
length(attr(d2, "format_unified"))
length(attr(d3, "format_unified"))

d <- cstidy::generate_test_data()
cstidy::set_csfmt_rts_data_v2(d)
ncol(d)

# Take the old class off first. set_csfmt_rts_data_v3() on a table that is
# still csfmt_rts_data_v2 leaves it a csfmt_rts_data_v2.
cstidy::remove_class_csfmt_rts_data(d)
cstidy::set_csfmt_rts_data_v3(d)
ncol(d)
class(d)
```

## v3 is weekly only

v3 fills in the time columns from `isoyearweek` and from nothing else. Weekly
data comes out complete.

A daily table keeps its `date` values and gets NA in every other time column.
Data that is not weekly stays on v2.

```{r}
weekly <- data.table::data.table(
  isoyearweek = c("2020-34", "2020-35"),
  location_code = "nation_nor",
  deaths_n = c(1L, 2L)
)
cstidy::set_csfmt_rts_data_v3(weekly)
weekly[, .(isoyearweek, isoyear, isoweek, season, seasonweek, date)]

daily <- data.table::data.table(
  granularity_time = "date",
  date = as.Date(c("2020-08-17", "2020-08-18")),
  location_code = "nation_nor",
  deaths_n = c(1L, 2L)
)
cstidy::set_csfmt_rts_data_v3(daily)
daily[, .(date, isoyear, isoweek, isoyearweek, season, seasonweek)]
```

## csdb cannot check a v3 table, though it can store one

`csdb` ships table validators for `csfmt_rts_data_v1` and `csfmt_rts_data_v2`,
and nothing for v3, so a v3 column set fails both. That does not stop you
storing it. The validator is an ordinary argument to `csdb::DBTable_v9$new()`:
pass `validator_field_types_blank()`, or a function of your own, and the table
writes. What you give up is the column check on the way in, not the storage.

# Where cstidy sits

cstidy imports `csdata` for Norwegian geography and `cstime` for time
conversions. It feeds `csalert`, whose `ens_collapse(heal = TRUE)` returns a
`csfmt_rts_data_v3`.

# Where to read next

- `vignette("csfmt_rts_data_v2", package = "cstidy")` — the column-by-column
  reference for v2.
- `vignette("benchmarks", package = "cstidy")` — timings.
- `?set_csfmt_rts_data_v3` — the v3 reference. No vignette covers v3 yet.

Everything below documents `csfmt_rts_data_v2`.

# csfmt_rts_data_v2

`csfmt_rts_data_v2` (`vignette("csfmt_rts_data_v2", package = "cstidy")`) is the Core Surveillance data format for real-time surveillance of infectious diseases.

```{r}
d <- cstidy::generate_test_data()
cstidy::set_csfmt_rts_data_v2(d)

# Looking at the dataset
d[]
```

## Smart assignment

`csfmt_rts_data_v2` supports smart assignment for time and geography. When the **bold** variables below are set with `:=`, the associated variables are automatically derived.

**location_code**:

- granularity_geo
- country_iso3

**isoyear**:

- granularity_time
- isoweek
- isoyearweek
- season
- seasonweek
- calyear
- calmonth
- calyearmonth
- date

**isoyearweek**:

- granularity_time
- isoyear
- isoweek
- season
- seasonweek
- calyear
- calmonth
- calyearmonth
- date

**date**:

- granularity_time
- isoyear
- isoweek
- isoyearweek
- season
- seasonweek
- calyear
- calmonth
- calyearmonth

```{r}
d <- cstidy::generate_test_data()[1:5]
cstidy::set_csfmt_rts_data_v2(d)

# Looking at the dataset
d[]

# Smart assignment of time columns (note how granularity_time, isoyear, isoyearweek, date all change)
d[1,isoyearweek := "2021-01"]
d

# Smart assignment of time columns (note how granularity_time, isoyear, isoyearweek, date all change)
d[2,isoyear := 2019]
d

# Smart assignment of time columns (note how granularity_time, isoyear, isoyearweek, date all change)
d[4:5,date := as.Date("2020-01-01")]
d

# Smart assignment fails when multiple time columns are set
d[1,c("isoyear","isoyearweek") := .(2021,"2021-01")]
d

# Smart assignment of geo columns
d[1,c("location_code") := .("norge")]
d

# Collapsing down to different levels, and healing the dataset 
# (so that it can be worked on further with regards to real time surveillance)
d[, .(deaths_n = sum(deaths_n), location_code = "norge"), keyby=.(granularity_time)] %>%
  cstidy::set_csfmt_rts_data_v2(create_unified_columns = FALSE) %>%
  print()

# Collapsing to different levels, and removing the class csfmt_rts_data_v2 because
# it is going to be used in new output/analyses
d[, .(deaths_n = sum(deaths_n), location_code = "norge"), keyby=.(granularity_time)] %>%
  cstidy::remove_class_csfmt_rts_data() %>%
  print()
```

## Summary

`summary()` gives a concise overview of the data structure.

```{r}
cstidy::generate_test_data() %>%
  cstidy::set_csfmt_rts_data_v2() %>%
  summary()
```

## Identifying the data structure of one column

`cstidy::identify_data_structure()` inspects a single column and returns a plottable object.

```{r}
cstidy::generate_test_data() %>%
  cstidy::set_csfmt_rts_data_v2() %>%
  cstidy::identify_data_structure("deaths_n") %>%
  plot()
```
